MSstatsPTM statistical relative quantification of post-translational modifications in global proteomics experiments

### 2021-10-27 main/MSstatsPTM_Main.tex
Updated figures and sample size sections

The proposed statistical framework allows for design of PTM experiments in terms of sample size calculation and power analysis. 
%Sample size calculation based on linear models has been described for general applications~\cite{kutner_etal_04a} and specifically for protein significance analysis~\cite{oberg_vitek_09a}. 
Sample size calculation takes as input a) $q$, the desired false discovery rate, b) $\beta$, the average Type II error rate, c) $\Delta$, the minimal log-fold change in adjusted PTM abundance that we would like to detect, d) $m_0 / (m_0 + m_1)$, the fraction of truly differentially modified PTM sites in the comparison, and e) $\sigma_{\gamma^{\ast}}^{2}$ and $\sigma_{\gamma}^{2}$, the anticipated variances associated to modified and unmodified peptide features, respectively. The variances can be derived based on the dataset being analyzed, assuming similar quantitative properties and variations. With these values and a user-specified number of conditions, the corresponding number of technical replicates per condition can then be derived, as described in~\cite{kutner_etal_04a}. Given the above quantities, the minimal number of replicates $J$ is determined by the variance of the estimated log-fold change $\mathrm{SE}^{2}(\hat{\Delta})$ as
\[
\mathrm{SE}^{2}(\hat{\Delta}) = \left[ \frac{2}{J} \left( \hat{\sigma}_{\gamma^{\ast}}^{2} + \hat{\sigma}_{\gamma}^{2} \right) \right]
\leq \left( \frac{\Delta}{t_{1-\beta, df} + t_{1-\alpha /2, df}} \right)^{2},
\]
where 
\[
\alpha = (1 - \beta) \cdot \frac{q}{1 + (1-q) \cdot m_0 / m_1},
\]
and $t_{1-\beta, df}$ and $t_{1-\alpha /2, df}$ are the $100(1-\beta)^{\text{th}}$ and the $100(1-\alpha /2)^{\text{th}}$ percentiles of the $t$ distribution, with $df = I(J-1)$ degrees of freedom in balanced designs. More details can be found in~\cite{oberg_vitek_09a}.

@@

\begin{align*}
H_{0}: \Delta = (\mu_{i}^{\ast} - \mu_{i'}^{\ast}) - (\mu_{i} - \mu_{i'}) &= 0 \\
H_{a}: \Delta = (\mu_{i}^{\ast} - \mu_{i'}^{\ast}) - (\mu_{i} - \mu_{i'}) &\neq 0
\end{align*}

Model inference is done in the same way to how MSstats~\cite{choi_etal_14a} and MSstatsTMT~\cite{Huang:2020} target the global protein. In cases where the design in unbalanced, a linear mixed effects model is fit, which takes into account all potential sources of variation, and restricted maximum likelihood (RML) is used to estimate the parameters of each model. Once the parameters are estimated, we can combine the models using modified versions of the formulas in Section \ref{sec:adjustment_summary}.

The log-fold change in the adjusted PTM abundance, $\Delta$, is now estimated by RML, and Equation (2) becomes 
\[
\hat{\Delta} = (\hat{\mu}_{RML_i}^{\ast} - \hat{\mu}_{RML_{i'}}^{\ast}) - (\hat{\mu}_{RML_i} - \hat{\mu}_{RML_{i'}})
\]

@@

After using the converters, the next step is peptide/protein summarization and missing value imputation. In the methods sections it was assumed that all the input data was summarized; in the implementation we need to run a specific summarization step to make this happen. For the modified run, the package summarizes features, PSMs, which include the same modification together. Features with multiple modification sites are not included with single site features and are summarized separately. For the global profiling run all unmodified features from the same protein are summarized together, up to the protein level. Additionally, the summarization includes global median normalization and normalizes between MS runs. The package uses an AFT model, as mentioned in Section \ref{sec:missing_imp}, to impute missing values, although this step is optional.

@@

\subsubsection*{Power analysis shows more replicates are needed in the PTM profiling run in cases of unequal variance.}

@@

\begin{figure}[ht]
\centering
\includegraphics[scale=.8]{images/fig1.png}
\caption{\todo{Meena?: Can you write about how MS for PTMs is run?}}
\label{fig:workflow}
\end{figure}

\begin{figure}[ht]
\centering
\includegraphics[scale=.8]{images/fig3.png}
\caption{
Data structure of a typical PTM experiment and goals of PTM characterization. (a) Schematic data representation, in a simplified case of two conditions and two replicate runs. Each PTM site is modeled and characterized separately, where a PTM is quantified with multiple spectral features (boxes), distinguished by different charge states of a peptide. The feature intensities are viewed as repeated measurements of the underlying abundance of the PTM, where the abundance in Condition i is denoted by i. Features corresponding to unmodified peptides are considered together to perform adjustment with respect to protein abundance, where the protein abundance in Condition i is denoted by i*. Peptides can be fully cleaved (solid lines) and/or partially cleaved (dashed lines). Some spectral features can be missing. (b) PTM relative quantification by statistical inference, which makes use of the feature intensities to infer the underlying PTM abundance and protein abundance with an estimate of associated uncertainty. (c) Model-based testing for differential PTM abundance, which corrects for the underlying protein abundance with a cost of increased uncertainty about the estimate of difference between conditions.}
\label{fig:data-structure}
\end{figure}

\begin{figure}[ht]
\centering
\includegraphics[scale=.8]{images/MSstatsPTM_design.png}
\caption{The workflow of MSstatsPTM and how it fits into the experimental pipeline. MSstatsPTM's workflow starts after modified and unmodified peptide quantification. First tool specific pre-processing is done, this includes modification site identification, general data cleaning, and formatting the data into the format needed for the package. The next step is feature level summarization, which summarizes features up to the modification level for the PTM data, and the protein level for the protein data. In the final step a model is fit to identify differential PTMs across condtions.}
\label{fig:msstatsptm_design}
\end{figure}

>>>>

The proposed statistical framework allows for design of PTM experiments in terms of sample size calculation and power analysis. 
%Sample size calculation based on linear models has been described for general applications~\cite{kutner_etal_04a} and specifically for protein significance analysis~\cite{oberg_vitek_09a}. 
Sample size calculation takes as input a) $q$, the desired false discovery rate, b) $\beta$, the average Type II error rate, c) $\Delta$, the minimal log-fold change in adjusted PTM abundance that we would like to detect, d) $m_0 / (m_0 + m_1)$, the fraction of truly differentially modified PTM sites in the comparison, and e) $\sigma_{\gamma^{\ast}}^{2}$ and $\sigma_{\gamma}^{2}$, the anticipated variances associated to modified and unmodified peptide features, respectively. The variances can be derived based on the dataset being analyzed, assuming similar quantitative properties and variations. With these values and a user-specified number of conditions, the corresponding number of technical replicates per condition can then be derived, as described in~\cite{kutner_etal_04a}. Given the above quantities, the minimal number of replicates $J$ is determined by the variance of the estimated log-fold change $\mathrm{SE}^{2}(\hat{\Delta})$ as

\begin{equation}
\mathrm{SE}^{2}(\hat{\Delta}) = \left[ \frac{2}{J} \left( \hat{\sigma}_{\gamma^{\ast}}^{2} + \hat{\sigma}_{\gamma}^{2} \right) \right]
\leq \left( \frac{\Delta}{t_{1-\beta, df} + t_{1-\alpha /2, df}} \right)^{2}
\end{equation}
where 
\begin{equation}
\alpha = (1 - \beta) \cdot \frac{q}{1 + (1-q) \cdot m_0 / m_1}
\end{equation}

and $t_{1-\beta, df}$ and $t_{1-\alpha /2, df}$ are the $100(1-\beta)^{\text{th}}$ and the $100(1-\alpha /2)^{\text{th}}$ percentiles of the $t$ distribution, with $df = I(J-1)$ degrees of freedom in balanced designs.

Solving for J, the number of biological replicates per group is calculated as

\begin{equation}
J \geq \frac{(2\hat{\sigma}_{\gamma^{\ast}}^{2} + 2\hat{\sigma}_{\gamma}^{2})(t_{1-\beta, df} + t_{1-\alpha /2, df})^2}{\Delta^2}
\end{equation}

More details on sample size calculation can be found in~\cite{oberg_vitek_09a}.

@@

Model inference is done in the same way to how MSstats~\cite{Choi:2014} and MSstatsTMT~\cite{Huang:2020} target the global protein. In cases where the design in unbalanced, a linear mixed effects model is fit, which takes into account all potential sources of variation, and restricted maximum likelihood (RML) is used to estimate the parameters of each model. Once the parameters are estimated, we can combine the models using modified versions of the formulas in Section \ref{sec:adjustment_summary}.

The log-fold change in the adjusted PTM abundance, $\Delta$, is now estimated by RML, and Equation (2) becomes 
\begin{equation}
\hat{\Delta} = (\hat{\mu}_{RML_i}^{\ast} - \hat{\mu}_{RML_{i'}}^{\ast}) - (\hat{\mu}_{RML_i} - \hat{\mu}_{RML_{i'}})
\end{equation}

@@

After using the converters, the next step is peptide/protein summarization and missing value imputation. In the methods sections it was assumed that all the input data was summarized; in the implementation we need to run a specific summarization step to make this happen. For the modified run, the package summarizes features, PSMs, which include the same modification together. Features with multiple modification sites are not included with single site features and are summarized separately. For the global profiling run all unmodified features from the same protein are summarized together, up to the protein level. Additionally, the summarization includes global median normalization and normalizes between MS runs. The package uses an AFT model to impute missing values, although this step is optional.

@@

\subsubsection*{Noisy PTM measurements benefit from additional biological replicates}

@@

\begin{figure}[ht]
\centering
\includegraphics[scale=.65]{images/fig3.png}
\caption{
Data structure of a typical PTM experiment and goals of PTM characterization. (a) Schematic data representation, in a simplified case of two conditions and two replicate runs. Each PTM site is modeled and characterized separately, where a PTM is quantified with multiple spectral features (boxes), distinguished by different charge states of a peptide. The feature intensities are viewed as repeated measurements of the underlying abundance of the PTM, where the abundance in Condition i is denoted by i. Features corresponding to unmodified peptides are considered together to perform adjustment with respect to protein abundance, where the protein abundance in Condition i is denoted by i*. Peptides can be fully cleaved (solid lines) and/or partially cleaved (dashed lines). Some spectral features can be missing. (b) PTM relative quantification by statistical inference, which makes use of the feature intensities to infer the underlying PTM abundance and protein abundance with an estimate of associated uncertainty. (c) Model-based testing for differential PTM abundance, which corrects for the underlying protein abundance with a cost of increased uncertainty about the estimate of difference between conditions.}
\label{fig:data-structure}
\end{figure}

\begin{figure}[ht]
\centering
\includegraphics[scale=.8]{images/MSstatsPTM_design.png}
\caption{The workflow of MSstatsPTM and how it fits into the experimental pipeline. MSstatsPTM's workflow starts after modified and unmodified peptide quantification. First tool specific pre-processing is done, this includes modification site identification, general data cleaning, and formatting the data into the format needed for the package. The next step is feature level summarization, which summarizes features up to the modification level for the PTM data, and the protein level for the protein data. In the final step a model is fit to identify differential PTMs and unmodified proteins across conditions and the PTM model is adjusted for changes in the unmodified protein.}
\label{fig:msstatsptm_design}
\end{figure}
